Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-14 ind_files/05_unification_discussion.tex
Discuss implications for effective population size

More precisely, the formalism studied in this thesis predicts a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection'. Noise-induced selection is only seen in finite populations and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models illustrate that evolution in finite population can proceed in the direction opposite to that predicted by the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{parsons_consequences_2010, houchmandzadeh_fluctuation_2015, behar_fluctuations-induced_2016, constable_demographic_2016,mcleod_social_2019}. Most recently, a study showed that in a very broad class of competition models, noise-induced selection in finite populations can act in opposition to the direction of natural selection predicted in the infinite population limit~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$.  Indeed, in ecological models, previous work has numerically shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of our birth-death formalism, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by reducing the birth rate through arresting the cell cycle, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent paper that uses certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration to arrive at a formalism mathematically equivalent to our equations~\citep{kuosmanen_turnover_2022}. The equivalent of my stochastic equations has also recently been derived for quantitative traits from a very different starting point\footnote{See Section \ref{sec_disc_field_eqns} above for a detailed discussion on how the field equation approach I used in \ref{chap_infD_processes} for quantitative traits is related to~\cite{week_white_2021}'s SPDE approach} using the theory of measure-valued branching processes~\citep{week_white_2021} - Equations (21b) and (21c) in \cite{week_white_2021} are exactly the $m \to \infty$ version of our equations for changes in the mean value of a type-level quantity and changes in the variance of a type-level quantity respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied.

On the practical side, the existence of noise-induced selection implies that simulation studies working with evolutionary individual-based or agent-based models should be careful about whether interaction effects are incorporated into birth rates or death rates since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection, thus potentially biasing results~\citep{mcleod_social_2019,kuosmanen_turnover_2022}. Being mindful of noise-induced selection is also important for applied fields like conservation and population management which regularly deal with small populations. For example, when trying to increase the population of a hypothetical desired species in a multispecies community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which leads to positive noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species.~\cite{raatz_promoting_2023} have recently used a similar birth-death framework to study cancer treatment and thus provide a concrete example of the consequences of this asymmetry between changing birth rates and death rates: Due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells in their model, with obvious implications for optimal treatment strategies. My results also indicate that measuring the growth rate of populations is not, in general, sufficient for accurate prediction/inference of future trajectories of the relative abundance of a species (or phenotype, allele, etc.) from empirical data even in completely controlled environments. The growth rate $w_i = b^{\textrm{(ind)}}_{i} - d^{\textrm{(ind)}}_{i}$ of a species $i$ only specifies the difference between its per-capita birth and death rates. In contrast, the complete stochastic dynamics also depend on the total turnover $\tau_i = b^{\textrm{(ind)}}_{i} + d^{\textrm{(ind)}}_{i}$ (\emph{i.e.} the sum of the per-capita birth and death rates).

>>>>

More precisely, the formalism studied in this thesis predicts a directional evolutionary force acting on variation in per-capita turnover rate $\tau$ that I call `noise-induced selection'. Noise-induced selection is only seen in finite populations and arises due to different types of individuals in the population experiencing a different number of stochastic events (birth and death) in a given time interval. Several specific finite population models illustrate that evolution in finite population can proceed in a direction different from that predicted by the infinite population limit~\citep{parsons_consequences_2010, houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, behar_fluctuations-induced_2016, constable_demographic_2016,abu_awad_effects_2018,parsons_pathogen_2018,mcleod_social_2019}. Indeed, in some stochastic models, the outcome of finite population models can be exactly opposite to that of the infinite population limit, a phenomenon sometimes referred to as `reversing the direction of deterministic selection'~\citep{constable_demographic_2016,mcleod_social_2019}. Most recently, a study showed that in a very broad class of competition models, noise-induced selection in finite populations can act in opposition to the direction of natural selection predicted in the infinite population limit~\citep{mazzolini_universality_2022}. My thesis generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}.

The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$.  Indeed, in ecological models, previous work has numerically shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of our birth-death formalism, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by reducing the birth rate through arresting the cell cycle, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent paper that uses certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration to independently arrive at a formalism mathematically equivalent to our equations~\citep{kuosmanen_turnover_2022}. The equivalent of my stochastic equations has also recently been derived for quantitative traits from a very different starting point\footnote{See Section \ref{sec_disc_field_eqns} above for a detailed discussion on how the field equation approach I used in \ref{chap_infD_processes} for quantitative traits is related to~\cite{week_white_2021}'s SPDE approach} using the theory of measure-valued branching processes~\citep{week_white_2021} - Equations (21b) and (21c) in \cite{week_white_2021} are exactly the $m \to \infty$ version of our equations for changes in the mean value of a type-level quantity and changes in the variance of a type-level quantity respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied.

On the practical side, the existence of noise-induced selection implies that simulation studies working with evolutionary individual-based or agent-based models should be careful about whether interaction effects are incorporated into birth rates or death rates since this seemingly arbitrary choice can have unintended consequences due to noise-induced selection, thus potentially biasing results~\citep{mcleod_social_2019,kuosmanen_turnover_2022}. Being mindful of noise-induced selection is also important for applied fields like conservation and population management which regularly deal with small populations. For example, when trying to increase the population of a hypothetical desired species in a multispecies community, increasing the birth rate is \emph{not} equivalent to reducing the death rate even though both result in an increase in the Malthusian fitness (growth rate) $w_i$. Decreasing the death rate leads to a decrease in $\tau_i$, which leads to positive noise-induced selection, whereas increasing the birth rate leads to an increase in $\tau_i$, which leads to noise-induced selection acting to reduce the abundance of the focal species from the community. If the total community size is small, increasing the birth rate of a species can thus lead to noise-induced selection completely eliminating the focal species from the community despite the fact that we \emph{increased} the growth rate of this species.~\cite{raatz_promoting_2023} have recently used a similar birth-death framework to study cancer treatment and thus provide a concrete example of the consequences of this asymmetry between changing birth rates and death rates: Due to the presence of noise-induced selection, the potential of a tumorous growth to adapt to treatments and experience evolutionary rescue depends inversely on the per-capita turnover $\tau_i$ of the constituent cancer cells in their model, with obvious implications for optimal treatment strategies. My results also indicate that measuring the growth rate of populations is not, in general, sufficient for accurate prediction/inference of future trajectories of the relative abundance of a species (or phenotype, allele, etc.) from empirical data even in completely controlled environments. The growth rate $w_i = b^{\textrm{(ind)}}_{i} - d^{\textrm{(ind)}}_{i}$ of a species $i$ only specifies the difference between its per-capita birth and death rates. In contrast, the complete stochastic dynamics also depend on the total turnover $\tau_i = b^{\textrm{(ind)}}_{i} + d^{\textrm{(ind)}}_{i}$ (\emph{i.e.} the sum of the per-capita birth and death rates). Lastly, noise-induced selection is particular to fluctuating populations and does not occur in models with fixed population sizes\footnote{If $N_K$ is a constant, the map $x_i \to x_i/N_K$ becomes a linear map and we thus no longer need It\^o's formula to move from densities to frequencies; Thus, equation \eqref{nD_FPE} directly provides the complete dynamics of the system in frequency space: Note that the directional terms here depend only on $\mathbf{A}^-$, which in turn depends only on the fitness $w_i$ and the mutation terms, and this system thus has no noise-induced selection} such as the neutral Wright-Fisher or Moran models, suggesting that working with such constant population frameworks is not sufficient to accurately capture the dynamics of real populations. Some previous theoretical studies have pointed out that approximating the population size of natural populations (which fluctuates) through a constant `effective population size' obtained as the harmonic mean of population size over is not always valid~\citep{sjodin_meaning_2005,abu_awad_effects_2018,kuosmanen_turnover_2022}. Recent experimental evolution studies have also directly shown that the harmonic mean of population size cannot always capture evolutionary population dynamics in fluctuating populations~\citep{chavhan_larger_2019}. The fact that noise-induced selection is only seen in fluctuating populations further underscores the general message from these studies - Approximating fluctuating populations via a constant effective population size may inadvertently remove important evolutionary properties of the systems under study.
